---
title: "R Workshop Day 5: Troubleshooting and Some Common Problems"
author: "Tuba Sendinc"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_notebook:
    number_sections: true
---


# Unmatched parantheses, curly braces, square brackets, or quotes

This is a very common error type. 

```{r}
## missing a closing parenthesis:
mean(c(1, 7, 13)
mean(c(1, 7, 13)) #corrected

## missing a curly brace before "else":
x <- 7 
if(x > 10) {
  print("x > 10")
   else {
  print("x <= 10")
 }

#corrected:
if(x > 10) {
  print("x > 10")
 } else {
  print("x <= 10")
   }
  
## missing a square bracket:
summary(ggplot2::diamonds[, "price")
summary(ggplot2::diamonds[, "price"] #corrected

## missing a closing quote in 2nd element:
group <- c("Group 1", "Group 2) 
group
group <- c("Group 1", "Group 2")  #corrected

```

If you are using RStudio, these types of mistakes are less likely because RStudio automatically write a full parenthesis, curly brace, square bracket or quotation mark when you open one.

# Not installing or loading a function

```{r}
## install package:
install.packages("ggplot2") #when installing a package, package name should be in quotation marks

## load package:
library(ggplot2) # when loading a package, package name should not be in quotation marks
```

# Typos 

```{r}
# typo in function name:
maen(c(1, 7, 13))

# corrected:
mean(c(1, 7, 13))

# typo in variable name:
summary(ggplot2::diamonds[, "detph"])

#corrected:
summary(ggplot2::diamonds[, "depth"])

# typo in dataset name:
data <- data.frame(
  x = rnorm(10),
  y = rnorm(10)
)
summary(dta[, 2])

#corrected:
summary(data[, 2])

# typo in object name:
test <- c(1, 7, 13)
mean(tset)

#corrected:
mean(test)

# typo in package name:
library("tydiverse")

#corrected:
library("tidyverse")
```

# Using inappropriate analysis for different data types 

Knowing our data type is very important because it dictates what we can do with our data. 

You cannot compute the mean of a character or factor vector.
```{r}
gender <- factor(c("female", "female", "male", "female", "male"))

mean(gender) #does not work
```

Instead, we should tabulate a character vector:
```{r}
table(gender) #3 females, 2 males!

prop.table(table(gender)) #this is how we look at proportions of each cell!

barplot(table(gender)) #this is how we visually display the distribution!
```

Mostly, it is not useful to create a barplot of a quantitative continuous vector, because frequency for each value will be mostly 1 and we cannot learn much about our data.
```{r}
barplot(table(rnorm(100)))
```

Instead, we can look at the summary statistics!
```{r}
x <- rnorm(100)
summary(x)
```

Or, we can draw a histogram or a box plot!
```{r}
par(mfrow = c(1, 2)) #combine two plots
hist(x)
boxplot(x)
```

# Not setting the correct working directory!!!

We should always know which folder is used as our working directory. getwd() function helps us with that. I recommend creating a separate folder for each assignment, and setting a separate working directory in each R script as to avoid these problems.
```{r}
getwd() #finding out the working directory
```

## How to Google "R Stuff"

Finding solutions for coding related problems can be challenging in the beginning. R's name also makes this challenging because it is a simple letter and it makes a google search harder. 

# Use quotation marks in Google search

When you are searching for a solution, try to add "in R" expression to your search. By doing this, you are more likely to get results relevant to your search. 
For instance, instead of searching for the following:

How to replace a character string r

Search for the following:
How to replace a character string "in R"

# Use package names in your Google search

When your problem involves a package, try to add it in your Google search, as it increases your chances of finding relevant results.

For instance, 

How to add axes labels in ggplot

Even better:

How to add axes labels "in ggplot"

# When you cannot find an answer...

Sometimes we run into issues that have nothing to do with our code. When this happens, turning R off and on can help! Save the information you need to save, and turn it off. Then turn it on again, clean your global environment, and make a fresh start.




