---
title: "Exercise X"
author: "Student Name"
date: "2023-08-15"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
#install packages 
#install.packages(c("knitr", "stargazer"))

#load in libraries
library(knitr)
library(stargazer)
```

In this illustration, I will use the mtcars dataset. The mtcars dataset is a built-in dataset in R that contains measurements on 11 different attributes for 32 different cars.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
data(mtcars)  #importing the mtcars dataset                          
```

## Part 1

Here is one way to create descriptive statistics table:
```{r}
descriptivestats <- data.frame(
    variablename = c("Miles/US Gallon", "Number of Cylinders", "Gross Horsepower"),
  mean  = c(mean(mtcars$mpg), mean(mtcars$cyl), mean(mtcars$hp)),
  sd = c(sd(mtcars$mpg), sd(mtcars$cyl), sd(mtcars$hp)),
  min  = c(min(mtcars$mpg), min(mtcars$cyl), min(mtcars$hp)),
  max = c(max(mtcars$mpg), max(mtcars$cyl), max(mtcars$hp))
)
kable(
  descriptivestats,
  col.names = c("Variable Name", "Mean", "St. Deviation", "Min", "Max"),
  digits = 2,
  caption = "Descriptive Statistics"
  )
```
This is the typical appearance of descriptive statistics tables in general.

## Part 2

Let's create the histogram of Miles/US Gallon variable!

```{r fig1, echo=FALSE, fig.cap="Distribution of Miles/US Gallon Variable"}
hist(mtcars$mpg,
     col='steelblue',
     main='',
     xlab='Miles/US Gallon',
     ylab='Frequency')
```

## Part 3

Finally, let's run some regressions and create a professional-looking regression results table!
```{r results='asis'}
#linear regression using lm() function
model1<-lm(mpg ~ gear, data = mtcars)
model2<-lm(mpg ~ carb, data = mtcars)
model3<-lm(mpg ~ carb + gear, data = mtcars)

#creating the regression table using stargazer() function
stargazer(model1, model2, model3,
title = "Regression Table with Stargazer",
dep.var.labels=c("Miles/US Gallon"),
covariate.labels=c("Number of Forward Gears","Number of Carburetors"),
label="tab-2",
column.labels = c("Model 1", "Model 2", "Model 3"),
model.numbers = FALSE,
header=FALSE)
```

